---
title: "Data Analyst Skills Test (HSM2)"
author: "Durand Sinclair"
date: "20/11/2021"
output: html
theme: yeti
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The following is Durand Sinclair's answers to take home test HSM2, which involves analysing simulated lung cancer data and generating visualisations. 


```{r prep, echo=FALSE, results='hide',warning=FALSE, message=FALSE}
# 0.  PREPARE ENVIRONMENT -------------------------------------------------
    
    # 0.1 Clear global environment ----
    rm(list = ls()) # Clears global environment
    gc() # Clears memory
    cat("\014") # Clear console

    # 0.2 Load libraries ----
    library(tidyverse) # Hadley Wickham's best libraries
    library(here) # File system navigation
    library(formattable) # Nicely formatted tables
    
    # 0.3 Name our colour palette ----
    customGreen0 = "#DeF7E9"
    customGreen = "#71CA97"
    customRed = "#ff7f7f"
    
# 1.  GET DATA ------------------------------------------------------------

    # Note: Synthetic cancer data comes from
    # https://simulacrum.healthdatainsight.org.uk/
    
    # 1.1 Download data if it's not on disk ----
    if (!dir.exists(here::here("data",
                               "simulacrum_release_v1.2.0.2017"
                               )
                    )
        ) {
        url <- paste0(
            "https://simulacrumhdi.s3.eu-west-2.amazonaws.com/",
            "simulacrum_release_v1.2.0.2017.zip"
        )
        temp_zipped <- tempfile()        
        temp_unzipped <- tempfile()
        download.file(url = url, 
                      destfile = temp_zipped,
                      timeout = max(120, getOption("timeout"))
        )
        # unzip(zipfile = temp_zipped, exdir = temp_unzipped)
        unzip(zipfile = temp_zipped, exdir = here::here("data"))
        unlink(c(temp_zipped, temp_unzipped))
    } 
    
    # 1.2 Load tables from disk ----
    patient_df_raw <- read_csv(here::here(
        "data",
        "simulacrum_release_v1.2.0.2017",
        "data",
        "sim_av_patient.csv"
    ))
    tumour_df_raw <- read.csv(here::here(
        "data",
        "simulacrum_release_v1.2.0.2017",
        "data",
        "sim_av_tumour.csv"
    ))
    vital_status_lookup <- read.csv(here::here(
        "data",
        "simulacrum_release_v1.2.0.2017",
        "data_dictionary_files",
        "zvitalstatus_lookup.csv"
    ))
    sex_lookup <- read.csv(here::here(
        "data",
        "simulacrum_release_v1.2.0.2017",
        "data_dictionary_files",
        "zsex_lookup.csv"
    )) 
    type_lookup <- read.csv(here::here(
        "data",
        "simulacrum_release_v1.2.0.2017",
        "data_dictionary_files",
        "zicd_lookup.csv"
    ))
    ethnicity_lookup <- read.csv(here::here(
        "data",
        "ethnicity_lookup.csv"
    ))
```

The patient data set contains 2,200,626 unique Patient IDs.
The tumour data set contains 2,200,626 unique Patient IDs, but 2,371,281 unique tumours, so some people got cancer more than once.

```{r eda, include=FALSE}
# 2) EXPLORE DATA  --------------------------------------------
    # Note: Lung Cancer patients are: SITE_ICD10_O2_3CHAR= C34

    
    # Let's see what the tables are
    glimpse(tumour_df_raw)
    skimr::skim(tumour_df_raw)
    length(unique(tumour_df_raw$PATIENTID))
    
    # The tumour dataset contains
    #   * 2.371,281 rows
    #   * 2,200,626 unique patients
    
    glimpse(patient_df_raw)
    length(unique(patient_df_raw$PATIENTID))
    
    # The patient dataset contains
    #   * 2,200,626 rows
    #   * 2,200,626 unique patients
```    
# Question 1

### a) What is the sex distribution of the patients?

```{r q1a, echo=FALSE, warning=FALSE, message=FALSE}
# Q1 --------------------------------------------------------------

    # Q1 a) What is the sex distribution of the patients? ----
    p <- patient_df_raw %>%
        inner_join(sex_lookup %>% select(ZSEXID, SHORTDESC), by = c("SEX" = "ZSEXID")) %>%
        select(Sex = SHORTDESC) %>%
        mutate(Sex = str_to_title(Sex)) %>%
        group_by(Sex) %>%
        count(name = "Total") %>%
        mutate(Total = accounting(Total, digits = 0))


    # 
    # # Tibble 
    # p %>% 
    #     tibble()
    
    # Formattable
    p %>%
        formattable(
            align = c("c", "r"),
            caption = "Sex distribution of all cancers",
            list(`Sex` = formatter("span", style = ~ style(color = "grey", font.weight = "bold")),
                 `Total` =  color_bar(customRed)
            ))

```

### b) Prepare a dataset with all lung cancer patients and their cause of death
Here are the first ten rows of that dataset.
```{r 1b, echo=FALSE, warning=FALSE}
# Q1 b) Prepare a dataset with all lung cancer patients and their cause of death ----

    # Note: Lung cancer = SITE_ICD10_O2_3CHAR= C34
    # Lookup table for causes of death is zicd_lookup.csv
    # Death classification is looked up in zvitalstatus.csv
    
    lung_death_df <-  tumour_df_raw %>%
        filter(SITE_ICD10_O2_3CHAR == "C34") %>%
        select(PATIENTID) %>%
        unique() %>%
        left_join(patient_df_raw, by = "PATIENTID") %>%
        filter(NEWVITALSTATUS == "D") %>%
        select(PATIENTID, DEATHCAUSECODE_UNDERLYING) %>%
        # 123,825
        separate(col = "DEATHCAUSECODE_UNDERLYING", 
                 into = c("a", "b", "c"), 
                 sep = ",",) %>%
        gather("key", "Code", a, b, c) %>%
        group_by(Code) %>%
        summarise(Total = n()) %>%
        inner_join(y = type_lookup, by = c("Code" = "ZICDID")) %>%
        select(Code, Description = SHORTDESC, Total) %>%
        arrange(desc(Total)) 
    
    lung_death_df %>%
        head(10) %>%
        mutate(Total=accounting(Total, digits=0)) %>%
        formattable(
            align = c("c", "l", "r"),
            caption = "Top 10 Causes Of Death For Lung Cancer Patients"
            )
```
### c) Calculate the mean survival time in days for lung cancer patients
The mean survival time for lung cancer patients is 299 days.
```{r q1c, echo=FALSE, messages=FALSE, warning=FALSE}
# Q1 c) Calculate the mean survival time in days for lung cancer patients ----
    survival_time <-  tumour_df_raw %>%
        filter(SITE_ICD10_O2_3CHAR == "C34") %>%
        select(PATIENTID, DIAGNOSISDATEBEST) %>%
        group_by(PATIENTID) %>%
        summarise(diagnosis = first(DIAGNOSISDATEBEST)) %>%
        unique() %>%
        left_join(patient_df_raw, by = "PATIENTID") %>%
        filter(NEWVITALSTATUS %in% c("D", "X")) %>%
        select(PATIENTID, diagnosis, VITALSTATUSDATE) %>%
        mutate(diagnosis = as.Date(diagnosis)) %>%
        mutate(days = VITALSTATUSDATE - diagnosis) %>%
        summarise(mean_survival_time_in_days = mean(days)) 

    survival_time
```
# Question 2

### a) Visualise the proportion of all ethnic groups among non-British male patients

```{r 2a, echo=FALSE}
# Q2 ----
# Q2 a) Visualise the proportion of all ethnic groups among non-British male patients ----
    ethnicity <- tumour_df_raw %>%
        select(PATIENTID) %>%
        unique() %>%
        left_join(patient_df_raw, by = "PATIENTID") %>%
        select(PATIENTID, SEX, ETHNICITY) %>%
        left_join(sex_lookup, by = c("SEX" = "ZSEXID")) %>%
        left_join(ethnicity_lookup, by = c("ETHNICITY" = "Code")) %>%
        select(PATIENTID, Sex = SHORTDESC, Ethnicity = Description) %>%
        filter(Sex == "MALE") %>%
        filter(Ethnicity != "White - British") 

    eth <- ethnicity %>%
        group_by(Ethnicity) %>%
        summarise(count = n()) %>%
        mutate(pct = formattable::percent(round(count / sum(count), 2))) %>%
        arrange(desc(count))
    
    ggplot(data = eth, mapping = aes(x = reorder(Ethnicity, pct), pct)) + 
        geom_bar(stat = "identity") + 
        scale_y_continuous(labels = scales::percent) +
        coord_flip() + 
        xlab("Ethnicity") + 
        ylab("Percent") + 
        ggtitle("Ethnicity of Male Cancer Patients") +
        theme_classic()
```

### b) Find an interesting or insightful aspect you can communicate visually.

```{r q2b, echo=FALSE}
# Q2 b) Explore the data to find an interesting or insightful aspect you can communicate visually. Be creative. -----


    # There's an idea that poorer people live shorter lives. 
    # We can see if this is true in this dataset because the subjects are 
    # grouped into socio-economic quintiles.
    # Let's see if it makes a difference to rates of cancer.

quintiles <- tumour_df_raw %>%
    select(QUINTILE_2015) %>%
    group_by(QUINTILE_2015) %>%
    summarise(observed = n()) %>%
    mutate(expected = nrow(tumour_df_raw)/5) %>%
    mutate(QUINTILE_2015 = as.factor(QUINTILE_2015))

ggplot(data = rev(quintiles), 
       mapping = aes(x = QUINTILE_2015,
                     y = observed)) +
    geom_point() +
    geom_hline(yintercept = quintiles$expected, colour="red") + 
    theme_light() + 
    #scale_y_discrete(limits=rev(levels(observed))) +
    xlab("Socio-economic Quintile") +
    ylab("Number of Patients") +
    scale_y_continuous(labels = comma) +
    scale_x_discrete(limits=rev(levels(quintiles$QUINTILE_2015))) +
    ggtitle("Simulated Cancer Patients Tend To Be Wealthier") 

```

# Bonus: RBCO Chart
The reason this job is being advertised is to recreate the Cancer Institute's RBCO charts. Here is a visualisation of the first chart in the series.

```{r rbco, echo=FALSE, warning=FALSE, message=FALSE}
# FUNCTION
chart_2 <- function(x, text_size = 8) {
    
    # Define text size
    outside_text <- text_size
    inside_text <- 4/14 * outside_text
    
    # Prepare data for charting
    df_segment_prep <- x %>%
        mutate(min_value = redmax - (yellowmax-redmax)) %>%
        mutate(max_value = yellowmax + (yellowmax-redmax)) %>%
        mutate(adjusted_dot = case_when(
            dot < redmax ~ (dot - (redmax -(yellowmax-redmax))) / 
                (redmax - (yellowmax-redmax)),
            dot < yellowmax ~ ((dot - redmax) / (yellowmax - redmax)) * 100 + 100,
            dot > yellowmax ~ ((dot - yellowmax) / 
                                   ((yellowmax + (yellowmax - redmax)) 
                                    - yellowmax)) * 100 + 200
        )) %>%
        mutate(my_group = factor(case_when(
            dot < redmax ~ "red",
            dot < yellowmax ~ "yellow",
            dot >= yellowmax ~ "green",
            TRUE ~ "Unknown"
        ))) %>%
        mutate(disease = as.character(paste0(disease, "(%) \n(N=", n, ")"))) %>%
        mutate(disease = factor(disease, levels = unique(disease))) %>%
        mutate(disease = forcats::fct_rev(disease)) %>%
        mutate(dot_label = as.character(dot))

    # Draw chart
    ggplot(df_segment_prep) +
        
        # Make the bars go horizontally
        coord_flip() +
        
        # Set up theme to adjust background elements
        theme(panel.grid.major = element_blank(), 
              panel.grid.minor = element_blank(),
              panel.background = element_blank(), 
              axis.line = element_line(colour = "grey"),
              axis.title = element_blank(),
              axis.text.x = element_blank(),
              axis.line.x = element_blank(),
              axis.ticks.x = element_blank(),
              legend.position = "none", 
              text = element_text(size = outside_text)
              ) +
        
        # Draw red, yellow & green lines
        geom_segment(aes(x = disease, xend = disease, y = 0, yend=100), 
                     color="red" ) +
        geom_segment(aes(x = disease, xend = disease, y = 100, yend=200), 
                     color = "orange") +
        geom_segment(aes(x = disease, xend = disease, y = 200, yend = 300), 
                     color = "forestgreen") +
        geom_point(aes(x = disease, y = adjusted_dot, color = my_group)) +
        
        # Ticks, dots and numbers
        geom_text(aes(
            x = disease, 
            y = 200,
            label = "|",
            alpha = 0.5
        ), size = inside_text) +
        geom_text(aes(
            x = disease,
            y = 200,
            label = notch,
            vjust = 2.0,
            alpha = 0.5
        ), size = inside_text) +
        geom_text(aes(
            x = disease,
            y = adjusted_dot,
            label = dot_label,
            vjust = -1.0,
            alpha = 0.5
        ), size = inside_text) +
        scale_colour_manual(values = c("forestgreen", "orange", "red")) +
        
        # Add a title at the top
        ggtitle("Performance Snapshot", 
                subtitle="Proportion of surgical episodes perfomed in hospitals above minimum suggested caseload")

}


# DATASET
df_02 <- tibble(
    disease = c("Lung", "Breast", "Gastric", "Oesophegeal", "Pancreatic", "Colon", "Rectal", "Kidney open approach", "Bladder", "Complex head and neck", "Primary Neurological", "Proportion of rectal cancer surgeries \nwith >= 12 nodes examined"),
    n = c(101, 733, 20, 25, 38, 542, 124, 51, 31, 217, 130, 31),
    redmax = c(78, 90, 53, 96, 88, 94, 67, 60, 10, 80, 96, 70),
    yellowmax = c(89, 92, 81, 98, 94, 97, 76, 69, 45, 84, 98, 85),
    dot = c(95, 91, 85, 100, 97, 99, 68, 69, 71, 84, 100, 87),
    notch = c(89, 92, 81, 98, 94, 97, 76, 69, 45, 84, 98, 85)
)

# VISUALISATION
chart_2(df_02, text_size = 8)
```
